---
title: "Introduction to life tables"
author: "Katie Paulson"
date: "March 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{introduction_to_life_tables}
  %\usepackage[UTF-8]{inputenc}
  %\VignetteDepends{demCore}
  %\VignetteDepends{hierarchyUtils}
---

```{r setup, include=FALSE}
library(data.table)
library(knitr)
library(kableExtra)
library(ggplot2)
devtools::load_all()
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## What is a life table?

A __life table__ is a table that includes information to describe the dying out
of a birth cohort. This can also be a synthetic birth cohort, in which case we
refer to it as a __period life table__.

Life tables are one of the most important devices in demography -- they have
been used since the 1600s! They can also be useful for other fields, because
they are generalizable to other discrete "time to event" data.

Typically, life tables have one row per age group, with columns representing
life table __metrics__, also known as __parameters__. The life table parameters
used in this package are $_nm_x$, $_na_x$, $_nq_x$, $_np_x$, $l_x$, $_nd_x$,
$e_x$, $_nL_x$, and $T_x$. In this notation, $x$ refers to age, and the metrics
apply to either the age $x$ directly or to the interval between ages $x$ and
$x+n$ where $n$ indicates the length of the interval, typically in years. We
often shorthand by removing the "n" from the notation, with interval length
implied.

Here's an example of a period life table, for males in Austria in 1992,
which is saved in the package as `austria_1992_lt` (source: Preston 2001):

```{r example_lt, echo = F}
data("austria_1992_lt")
dt <- austria_1992_lt[
  , .SD, .SDcols = c("age_start", "age_end", "deaths", "pop", "mx", "ax", "qx",
                     "px", "lx", "dx", "nLx", "Tx", "ex")
]
dt_rounded <- round(dt, 2)
setnames(dt_rounded, "age_start", "x")
setnames(dt_rounded, "age_end", "x+n")
kable(dt_rounded, format = "markdown")
```

## Life table metrics

For reference, the following is a list of life table metrics and their
definitions.

__$\mathbf{_nm_x}$:__ mortality rate between ages $x$ and $x+n$. Shorthand to
$m_x$ with implied interval width ($n$). Equals deaths divided by person-years
lived in the interval. Mid-year population is commonly used as an adequate
approximation of the person-years denominator.

__$\mathbf{_na_x}$:__ mean person-years lived between ages $x$ and $x+n$ for
those who die within the interval. Shorthand to $a_x$ with implied interval
width ($n$).  

__$\mathbf{_nq_x}$:__ probability of death between ages $x$ and $x+n$,
conditional on survival to age $x$. Shorthand to $q_x$ with implied interval
width ($n$). Equals deaths in the interval divided by survivors to $x$-th
birthday. Examples: $_5q_0$ = probability of death between birth and age $5$;
$_{45}q_{15}$ = probability of death between age $15$ and age $60$ conditional
on survival to age $15$.  

__$\mathbf{l_x}$:__ proportion of the cohort surviving to age $x$.  

__$\mathbf{e_x}$:__ life expectancy at age $x$ -- mean number of years lived
after $x$-th birthday by those surviving to age $x$. Life expectancy at birth
is $e_0$.  

__$\mathbf{_nL_x}$:__ total person-years lived between age $x$ and $x+n$.  

__$\mathbf{T_x}$:__ total person-years lived above age $x$.  

__$\mathbf{_nd_x}$:__ proportion of the cohort dying between ages $x$ and $x+n$.
Shorthand to $d_x$.  

__$\mathbf{_np_x}$:__ probability of survival between ages $x$ and $x+n$
conditional on survival to age $x$. Inverse of $qx$.

## Representing life tables graphically

We often reduce life tables to age patterns of log probability of death
($\text{log}(q_x)$) or to survival curves ($l_x$ over age), which can be easily
displayed and vetted in plots.

```{r plot_qx, echo = F}
ggplot(data = dt,
       aes(x = age_start,
           y = log(qx))) +
       geom_line(color = "magenta") +
       geom_point() +
       scale_x_continuous(breaks = c(0, 1, seq(5, 95, 5))) +
       theme_bw() +
       theme(axis.text.x = element_text(size = 6)) +
       ggtitle("log-qx over age")

ggplot(data = dt,
       aes(x = age_start,
           y = lx)) +
       geom_line(color = "magenta") +
       geom_point() +
       scale_x_continuous(breaks = c(0, 1, seq(5, 95, 5))) +
       theme_bw() +
       theme(axis.text.x = element_text(size = 6)) +
       ggtitle("lx over age (survival curve)")
```


## Calculations and relationships between life table metrics

The `demCore` package includes many utility functions for calculations
that leverage the mathematical relationships between life table metrics to
build out a complete life table. This section will provide details and examples
regarding the use of these functions and their underlying methods. We will
accomplish this by following along the example of building the example
life table above from death counts and population.

Note that this document and this package do not contain an exhaustive list of
relationships between metrics. Additionally, some equations presented rely on
assumptions and others are true relationships that are always valid. For more
details, see the Preston Demography textbook, from which many of these details
were drawn.

### mx

From raw death count and population data, the place to start with a life table
is $m_x$.

$$m_x = \frac{\text{deaths}}{\text{person-years}} \approx
\frac{\text{observed deaths}}{\text{mid-interval population}}$$

Let's load in our example data and calculate $m_x$:

```{r mx_calc}
data("austria_1992_lt")
dt <- austria_1992_lt[, c("age_start", "age_end", "deaths", "pop")]
dt[, mx := deaths / pop]
```

### ax

If we have $m_x$ and $q_x$ we can directly calculate $a_x$. However, we often
use $m_x$ and $a_x$ to get $q_x$ in the first place, and so have to make some
assumptions to get $a_x$. Empirical calculations of $a_x$ would require detailed
and accurate data on age of death in days (such as paired date of birth and date
of death), which is typically unavailable.

__Rule of thumb:__

One option is to assume all deaths occur in the middle of the interval, so
$a_x \approx n/2$. This assumption works well for most ages, but it doesn't work
as well for very young or very old where mortality can change rapidly over the
interval.

Another assumption we can make is that the age-specific death rate is constant
between $x$ and $x+n$. Under this assumption,
$$_na_x = n + \frac{1}{_nm_x} - \frac{n}{1- e^{-n  \cdot {_nm_x}}}.$$
The function `mx_to_ax` implements this assumption.

Using our example data, we get:

```{r ax_calc}
dt <- hierarchyUtils::gen_length(dt, col_stem = "age")
dt[, ax := mx_to_ax(mx = mx, age_length = age_length)]
```

Note that we can use `hierarchyUtils::gen_length` to add the `age_length` column
given `age_start` and `age_end`.

__1a0 and 4a1:__

Preston et al adapted an analysis first completed by Coale and Demeny (1983)
to derive a relationship between infant mortality rate ($_1m_0$) and under-5
$a_x$ values ($_1a_0$ and $_4a_1$). In the absence of reliable data to produce
$a_x$, these relationships can be used to predict $a_x$ from infant $m_x$:

|                     | Males       | Females  |
| ------------------- |-------------| ---------|
| 1a0:                |             |          |
| If 1m0 >= 0.107     | 0.330       | 0.350    |
| If 1m0 < 0.107      | 0.045 + 2.684 * 1m0 | 0.053 + 2.800 * 1m0 |
|                     |             |          |
| 4a1:                |             |          |
| If 1m0 >= 0.107     | 1.352       | 1.361    |
| If 1m0 < 0.107      | 1.651 - 2.816 * 1m0 | 1.522 - 1.518 * 1m0 |

Use the `gen_u5_ax` function to implement this method:

```{r ax_u5_calc, eval = F}
dt[, sex := "male"]
gen_u5_ax(dt, id_cols = c("age_start", "age_end", "sex"))
```

__Graduation method:__ One strategy for selecting $a_x$ values is based on the
level and slope of the $_nm_x$ function. Comparing two populations with the
same $_5m_60$, the population with more rapidly rising mortality rate with
respect to age will have deaths that are more concentrated in the later part of
the interval (higher $a_x$). Comparing two populations with the same slope in
$m_x$, the one with higher mortality rate will have more deaths at the beginning
of the interval (lower $a_x$).

To utilize this theory, we can implement iteration as described in the
Preston book, and originally proposed by Keyfitz (1966):
$$_na_x = \frac{\frac{-n}{24} {_nd_{x-n}} + \frac{n}{2} {_nd_x} +
\frac{n}{24} {_nd_{x+n}}}{_nd_x}$$
Where $d_x$ is derived from the conversion from $m_x$ to $q_x$. However, since
the $m_x$ to $q_x$ conversion requires $a_x$, this requires us to pick a
starting place for $a_x$ (like $n/2$), solve for $d_x$, solve for $a_x$, and so
on until convergence. Use `demCore::iterate_ax` to implement this method.

### qx

From $m_x$ and $a_x$, we can solve directly for $q_x$:

$$_nq_x = \frac{n \cdot {_nm_x}}{1 + (n - {_na_x}) \cdot {_nm_x}}$$
For the terminal age group, $q_x$ should be $1$ because all individuals
surviving to the terminal age group will die in that age group (probability of
death = $1$).

The `mx_ax_to_qx` combines the equation for $q_x$ and the requirement that
terminal $q_x$ equal one by setting $q_x = 1$ if `age_length = Inf`.

```{r qx_calc1, echo = T, results = "hide"}
dt[, qx := mx_ax_to_qx(mx = mx, ax = ax, age_length = age_length)]
```

Other functions that utilize this relationship but solve for different metrics
are `mx_qx_to_ax` and `qx_ax_to_mx`.

You can also solve for $q_x$ under the assumption of constant mortality rate
within an interval, which removes $a_x$ from the relationship:

$$_nq_x = 1 - e^{-n \cdot {_nm_x}}.$$
```{r qx_calc2, eval = F}
dt[, qx_compare := mx_to_qx(mx = mx, age_length = age_length)]
```

These two $q_x$ values are the same, because the implied $a_x$ in `mx_to_qx` is
equivalent to the $a_x$ we generate under the assumption in `mx_to_ax`.

### lx

To calculate the proportion of a cohort surviving to age $x$ ($l_x$), we set
$l_0 = 1$ (100% survive to birth), and recursively calculate:

$$l_{x+n} = l_x \cdot (1 - _nq_x)$$

or in words, the proportion surviving to age $x$ times the proportion of those
survivors who do not die between $x$ and $x+n$ is the proportion surviving to
age $x+n$.

Our `gen_lx_from_qx` function can perform this calculation:

```{r lx_calc}
gen_lx_from_qx(dt, id_cols = c("age_start", "age_end"))
```

### dx

Proportion of cohort dying between ages $x$ and $x+n$ ($d_x$) is $_nq_0$ to
start, then $_nd_x = l_x - l_{x+n}$ thereafter (difference between proportion
surviving to age $x$ and proportion surviving to age $x+n$).

To calculate $d_x$, use `gen_dx_from_lx`:

```{r dx_calc}
gen_dx_from_lx(dt, id_cols = c("age_start", "age_end"))
```

### nLx

The person-years lived between ages $x$ and $x+n$ ($_nL_x$) can be broken down
into:

* Person-years lived by those who _survive_ the interval = $n \cdot l_{x+n}$
* Person-years lived by those who _die_ during the interval = $_na_x \cdot {_nd_x}$

such that:

$$_nL_x = n \cdot l_{x+n} + _na_x \cdot {_nd_x}.$$

For the _terminal age group_:

$${_{\infty}L_x} = \text{person-years lived above age } x =
\frac{\text{person-years lived above age }x}{\text{deaths over age }x}
\cdot \text{deaths over age }x= \frac{l_x}{_{\infty}m_x}$$

Use the `gen_nLx` function to calculate with this method:

```{r nLx_calc}
gen_nLx(dt, id_cols = c("age_start", "age_end"))
```

### Tx

Next, use $_nL_x$ to get $T_x$:

$$T_x = \sum_{x}^{\infty} {_nL_x}.$$

```{r Tx_calc}
gen_Tx(dt, id_cols = c("age_start", "age_end"))
```

### ex

Life expectancy above age $x$ (mean person-years lived above age $x$) is equal
to the total person years over age $x$ divided by the persons surviving to age
$x$:

$$e_x = \frac{T_x}{lx}.$$

For the terminal age group, $a_x = e_x$ because everyone surviving to the
interval dies in the interval.

Calculate $e_x$ with `gen_ex`:

```{r ex_calc}
gen_ex(dt)
```

### Summary

One possible set of steps for calculating a complete period life table from 
deaths and mid-year population is:

1. Compute mx = deaths / population
2. Estimate <5 ax from mx using `gen_u5_ax`, and set ax over age 5 as n/2
3. Calculate qx directly using `mx_ax_to_qx`
4. Use ax iteration with `iterate_ax` to modify ax and qx values, improving
  ax over the naive n/2 values
5. Calculate px, lx, dx, nLx, Tx, and ex directly, in that order, from mx, ax, 
  and qx, with `lifetable` function. The `lifetable` function combines many
  of the functions described in this vignette for convenience.
6. In the terminal age group, ax is equal to ex. So, once ex is solved for,
  terminal ax can be replaced with that value.

### References

Preston Samuel H, Patrick H, Michel G. Demography: measuring and modeling
population processes. MA: Blackwell Publishing. 2001.

Coale AJ, Demeny P, Vaughan B. Regional model life tables and stable
populations: studies in population. Elsevier; 2013 Oct 22.

Keyfitz N. A life table that agrees with the data. Journal of the American
Statistical Association. 1966 Jun 1;61(314):305-12.



