---
title: "Introduction to life tables"
author: "Katie Paulson"
date: "3/6/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Title of your vignette}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
library(data.table)
library(knitr)
library(kableExtra)
library(ggplot2)
devtools::load_all()
knitr::opts_chunk$set(echo = TRUE)
```

## What is a life table?

A __life table__ is a table that includes information to describe the dying out
of a birth cohort. This can also be a synthetic birth cohort, in which case we
refer to it as a __period life table__.

Life tables are one of the most important devices in demography -- they have
been used since the 1600s! They can also be useful for other fields, because
they are generalizable to other discrete "time to event" data.

Typically, life tables have one row per age group, with columns representing
life table __metrics__, also known as __parameters__. The life table parameters
used in this package are $_nm_x$, $_na_x$, $_nq_x$, $_np_x$, $l_x$, $_nd_x$,
$e_x$, $_nL_x$, and $T_x$. In this notation, $x$ refers to age, and the metrics
apply to either the age $x$ directly or to the interval between ages $x$ and
$x+n$ where $n$ indicates the length of the interval, typically in years. We
often shorthand by removing the "n" from the notation, with interval length
implied.

Here's an example of a life table for Norweigian females in 2010 (from GBD 2019
results):

```{r example_lt, echo = F}
data("fNOR2010")
dt <- lifetable(fNOR2010, id_cols = c("age", "location"), terminal_age = 95)
dt <- dt[, .SD, .SDcols = c("age", "age_length", "deaths", "population",
                            "mx", "ax", "qx", "px", "lx", "dx", "nLx",
                            "Tx", "ex")]
dt_rounded <- round(dt, 2)
setnames(dt_rounded, "age", "x")
setnames(dt_rounded, "age_length", "n")
kable(dt_rounded, format = "markdown")
```

We also often reduce life tables to age patterns of log probability of death ($\text{log}(q_x)$) or to survival curves ($l_x$ over age), which can be
easily displayed and vetted in plots.

```{r plot_qx, echo = F}
ggplot(data = dt,
       aes(x = age,
           y = log(qx))) +
       geom_line(color = "magenta") +
       geom_point() +
       scale_x_continuous(breaks = c(0,1,seq(5,95,5))) +
       theme_bw() +
       ggtitle("NOR 2010 female life table: log-qx over age")

ggplot(data = dt,
       aes(x = age,
           y = lx)) +
       geom_line(color = "magenta") +
       geom_point() +
       scale_x_continuous(breaks = c(0,1,seq(5,95,5))) +
       theme_bw() +
       ggtitle("NOR 2010 female life table: lx over age (survival curve)")
```

## Life table metrics

For reference, the following is a list of life table metrics and their
definitions.

__$\mathbf{_nm_x}$:__ mortality rate between ages $x$ and $x+n$. Shorthand to
"mx" with implied interval width (n). Equals deaths divided by person-years
lived in the interval. Mid-year population is commonly used as an adequate
approximation of the person-years denominator.

__$\mathbf{_na_x}$:__ mean person-years lived between ages $x$ and $x+n$ for
those who die within the interval. Shorthand to "ax" with implied interval
width (n).  

__$\mathbf{_nq_x}$:__ probability of death between ages $x$ and $x+n$,
conditional on survival to age $x$. Shorthand to "qx" with implied interval
width (n). Equals deaths in the interval divided by survivors to x-th birthday.
Examples: 5q0 = probability of death between birth and age 5;
45q15 = probability of death between age 15 and age 60 conditional on survival
to age 15.  

__$\mathbf{l_x}$:__ proportion of the cohort surviving to age $x$.  

__$\mathbf{e_x}$:__ life expectancy at age $x$ -- mean number of years lived
after x-th birthday by those surviving to age $x$. Life expectancy at birth
is $e_0$.  

__$\mathbf{_nL_x}$:__ total person-years lived between age $x$ and $x+n$.  

__$\mathbf{T_x}$:__ total person-years lived above age $x$.  

__$\mathbf{_nd_x}$:__ proportion of the cohort dying between ages $x$ and $x+n$.
Shorthand to "dx."  

__$\mathbf{_np_x}$:__ probability of survival between ages $x$ and $x+n$
conditional on survival to age $x$. Inverse of $qx$.  

## Calculations and relationships between life table metrics

The `lifetableUtils` package includes many utility functions for calculations
that leverage the mathematical relationships between life table metrics to
build out a complete life table. This section will provide details and examples
regarding the use of these functions and their underlying methods. We will
accomplish this by following along the example of building the Norweigian
life table above from death counts and population.

Note that this document and this package do not conatain an exhaustive list of
relationships between metrics. Additionally, some equations presented rely on
assumptions and others are true relationships that are always valid. For more
details, see the Preston Demography textbook, from which many of these details
were drawn (cite here).

### mx

From raw death count and population data, the place to start with a life table
is mx.

$$m_x = \frac{\text{deaths}}{\text{person-years}} \approx \frac{\text{observed deaths}}{\text{mid-interval population}}$$
Let's load in our example Norweigian data and calculate mx:

```{r mx_calc}
data("fNOR2010")
dt <- fNOR2010[, c("age", "age_length", "deaths", "population")]
dt[, mx := deaths / population]
```

### ax

If we have mx and qx we can directly calculate ax. However, we often use mx
and ax to get qx in the first place, and so have to make some assumptions to
get ax. Empirical calculations of ax would require detailed and accurate data
on age of death in days (such as paired date of birth and date of death), which
is typically unavailable.

__Rule of thumb:__

One option is to assume all deaths occur in the middle of the interval, so
$ax \approx n/2$. This assumption works well for most ages, but it doesn't work
as well for very young or very old where mortality can change rapidly over the
interval.

Another assumption we can make is that the age-specific death rate is constant
between $x$ and $x+n$. Under this assumption,
$$_na_x = n + \frac{1}{_nm_x} - \frac{n}{1- e^{-n  \cdot {_nm_x}}}.$$
The function `mx_to_ax` implements this assumption.

Using our example data, we get:

```{r ax_calc}
dt[, ax := mx_to_ax(mx = mx, t = age_length)]
```


__1a0 and 4q1:__

Coale and Demeny (1983) produced empirically derived values for $_1a_0$ and
$_4a_1$:

|                     | Males       | Females  |
| ------------------- |-------------| ---------|
| 1a0:                |             |          |
| If 1m0 >= 0.107     | 0.330       | 0.350    |
| If 1m0 < 0.107      | 0.045 + 2.684 * 1m0 | 0.053 + 2.800 * 1m0 |
|                     |             |          |
| 4a1:                |             |          |
| If 1m0 >= 0.107     | 1.352       | 1.361    |
| If 1m0 < 0.107      | 1.651 - 2.816 * 1m0 | 1.522 - 1.518 * 1m0 |

__Graduation method:__ One strategy for selecting ax values is based on the
level and slope of the $_nm_x$ function. Comparing two populations with the
same 5m60, the population with more rapidly rising mortality rate with respect
to age will have deaths that are more concentrated in the later part of the
interval (higher ax). Comparing two populations with the same slope in mx, the
one with higher mortality rate will have more deaths at the beginning of the
interval (lower ax).

To utilize this theory, we can implement iteration as described in the
Preston book, and originally proposed by Keyfitz (1966):
$$_na_x = \frac{\frac{-n}{24} {_nd_{x-n}} + \frac{n}{2} {_nd_x} + \frac{n}{24} {_nd_{x+n}}}{_nd_x}$$
Where dx is derived from the conversion from mx to qx. However, since the
mx --> qx conversion requires ax, this requires us to pick a starting place for
ax (like n/2), solve for dx, solve for ax, and so on until convergence.

### qx

From mx and ax, we can solve directly for qx:

$$_nq_x = \frac{n \cdot {_nm_x}}{1 + (n - {_na_x}) \cdot {_nm_x}}$$
For the terminal age group, qx should be 1 because all individuals surviving to
the terminal age group will die in that age group (probability of death = 1).

```{r qx_calc1}
dt[, qx := mx_ax_to_qx(mx = mx, ax = ax, t = age_length)]

# set qx = 1 and recalculate ax for terminal age group
dt[age == max(dt$age), qx := 1]
dt[age == max(dt$age), ax := mx_qx_to_ax(mx = mx, qx = 1, t = 30)]
```

Other functions that utilize this relationship but solve for different metrics
are `mx_qx_to_ax` and `qx_ax_to_mx`.

You can also solve for qx under the assumption of contant mortality rate within
an interval, which removes ax from the relationship:

$$_nq_x = 1 - e^{-n \cdot {_nm_x}}.$$
```{r qx_calc2, eval = F}
dt[, qx_compare := mx_to_qx(mx = mx, t = age_length)]
```

These two qx values are the same, because the implied ax in `mx_to_qx` is
equivalent to the ax we generate under the assumption in `mx_to_ax`.

### lx

To calculate the proportion of a cohort surviving to age $x$ (lx), we set
$l_0 = 1$ (100% survive to birth), and recursively calculate:

$$l_{x+n} = l_x \cdot (1 - _nq_x)$$

or in words, the proportion surviving to age $x$ times the proportion of those
survivors who do not die between $x$ and $x+n$ is the proportion surviving to
age $x+n$.

Our `qx_to_lx` function can perform this calculation:

```{r lx_calc}
dt <- qx_to_lx(dt, id_cols = c("age"))
```

### dx

Proportion of cohort dying between ages $x$ and $x+n$ (dx) is $_nq_0$ to start,
then $_nd_x = l_x - l_{x+n}$ thereafter (difference between proportion surviving
to age $x$ and proportion surviving to age $x+n$).

To calculate dx, use `lx_to_dx`:

```{r dx_calc}
dt <- lx_to_dx(dt, id_cols = c("age"), terminal_age = 95)
```

### nLx

The person-years lived between ages $x$ and $x+n$ ($_nL_x$) can be broken down
into:

* Person-years lived by those who _survive_ the interval = $n \cdot l_{x+n}$
* Person-years lived by those who _die_ during the interval = $_na_x \cdot {_nd_x}$

such that:

$$_nL_x = n \cdot l_{x+n} + _na_x \cdot {_nd_x}.$$

For the _terminal age group_:

$${_{\infty}L_x} = \text{person-years lived above age } x =
\frac{\text{person-years lived above age }x}{\text{deaths over age }x}
\cdot \text{deaths over age }x= \frac{l_x}{_{\infty}m_x}$$

Use the `gen_nLx` function to calculate with this method:

```{r nLx_calc}
dt <- gen_nLx(dt, id_cols = c("age"), terminal_age = 95)
```

### Tx

Next, use $_nL_x$ to get $T_x$:

$$T_x = \sum_{x}^{\infty} {_nL_x}.$$

```{r Tx_calc}
dt <- gen_Tx(dt, id_cols = c("age"))
```

### ex

Life expectancy above age $x$ (mean person-years lived above age $x$) is equal
to the total person years over age $x$ divided by the persons surviving to age
$x$:

$$e_x = \frac{T_x}{lx}.$$

For the terminal age group, $a_x = e_x$ because everyone surviving to the
interval dies in the interval.

Calculate $e_x$ with `gen_ex`:

```{r ex_calc}
dt <- gen_ex(dt)
```

### Summary

The majority of these functions can be combined for convenience with the
`lifetable` function. `lifetable` requires at least two of $m_x$, $a_x$, and
$q_x$, but solves out the remaining life table metrics following the methods
and functions described here.

The resulting life table is the life table at the top of this document.
